Bayesian Hierarchical Model for estimating holocene prehistoric population development on the Swiss Plateau

### 2022-04-25 manuscript/paper.Rmd
first batch of incorporating Joes suggestions

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

library(here)
library(ggplot2)
library(ggmap)
library(sp)
library(sf)
library(ggsn)
library(rnaturalearth)
library(ggrepel)

@@

@@

my_basemap <- ggmap(swiss_background, darken = c(0.6, "white"))

my_scalebar <- scalebar(transform = T, dist_unit = "km", dist=50, model = "WGS84", st.dist = 0.025, st.size = 3, x.min = bbox_ch[1], x.max = bbox_ch[3]-0.2, y.min = bbox_ch[2]+0.15, y.max = bbox_ch[4], border.size = 0.5)
```

@@

<!-- JR: Might the following four paragraphs fit better in a background section ("Prehistoric demography")? -->
Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].

@@

<!-- JR: This paragraph also sounds like background to me. -->
Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data. Most notably are the recently published papers by Crema and Shoda ([@crema2021a]). However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step in the direction of a model-based, and thus scientific, analysis. However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series. Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented. Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.

@@

<!-- JR: Begin by defining, broadly, what Bayesian statistics is? -->
At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.). Our assumptions about the demographic development of the past must naturally be very conservative. In the logic of the basic Bayesian equation, these assumptions represent the prior (probability). In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian learning from data. It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction. Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.

@@

<!-- JR: Non-Europeanists (like me!) might appreciate a brief summary of the geography and prehistory of the Swiss plateau here. -->

Our working region consists of the Swiss Plateau, which on the one hand offers excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.

@@

<!-- JR: Would be nice to add rivers and lakes to the base map. -->

@@

```{r compdendroc14, fig.cap="Comparison of the scaled number of ^14^C and dendrodated sites over time in the dataset used."}
joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))

@@

<!-- JR: I don't understand what "scaled values" means here. -->

@@

@@

Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model. Preservation conditions again simply prevent their use in this study. Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.

<!-- JR: What preservation conditions? There are prehistoric burials and cemeteries in Switzerland, no? -->

@@

In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods <!-- JR: Can we gloss what autocorrelated means in plain language here, to make it clear that this is a trivial assumption? something like ", i.e. the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on." -->. In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time. This results in the following very simple formula:

@@

Thus, we can define the limits, the $max\_change\_rate$<!-- JR: no Greek letter? :( --> as an time independent variable, again without already specifying them with fixed values at this stage. The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively. Overall, this represents a hierarchical model that can be noted as follows:

@@

The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure \@ref(fig:gammadist)).

<!-- JR: A whole figure just to illustrate the gamma distribution seems like overkill. Could we add the prior distributions of other parameters? Or just get rid of it: you can always look up the shapes on Wikipedia :) -->

```{r gammadist, fig.cap="The gamma-distributed maximum growth rate prior of the model."}
plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
```

This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model. The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:

>>>>

```{r setup, echo = FALSE, message = FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

library(here)
library(ggplot2)
library(ggmap)
library(cowplot)
library(sp)
library(sf)
library(ggsn)
library(rnaturalearth)
library(ggrepel)

@@

switzerland_bb <- st_as_sfc(st_bbox(switzerland))

@@

rivers10 <- ne_download(scale = "large", type = 'rivers_lake_centerlines', category = 'physical') %>% st_as_sf()

rivers10eu <- ne_download(scale = "large", type = 'rivers_europe', category = 'physical') %>% st_as_sf()

lakes10 <- ne_download(scale = "large", type = 'lakes', category = 'physical') %>% st_as_sf()

lakes10eu <- ne_download(scale = "large", type = 'lakes_europe', category = 'physical') %>% st_as_sf()

my_basemap <- ggmap(swiss_background, darken = c(0.6, "white")) + coord_sf(default_crs = st_crs(4326)) + geom_sf(data = rivers10, color="lightblue4", inherit.aes = F)  + geom_sf(data = rivers10eu, color="lightblue4", inherit.aes = F)  + geom_sf(data = lakes10, fill="lightblue4", color=NA, inherit.aes = F)  + geom_sf(data = lakes10eu, fill="lightblue4", inherit.aes = F, color=NA)

my_scalebar <- scalebar(transform = T, dist_unit = "km", dist=50, model = "WGS84", st.dist = 0.025, st.size = 3, x.min = bbox_ch[1], x.max = bbox_ch[3]-0.2, y.min = bbox_ch[2]+0.15, y.max = bbox_ch[4], border.size = 0.5)

europe <- c(left = -12, bottom = 35, right = 30, top = 63) %>% as.vector() %>% +0.01

europe_overview <- get_stamenmap( bbox = europe, maptype = "toner-background", color="bw", zoom = 3)

europe_overview_map <- ggmap(europe_overview, darken = c(0.6, "white"))+ coord_sf(default_crs = st_crs(4326))
```

@@

<!-- JR: Might the following four paragraphs fit better in a background section ("Prehistoric demography")? -->

Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].

@@

<!-- JR: This paragraph also sounds like background to me. -->

Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data. Most notably are the recently published papers by Crema and Shoda ([@crema2021a]). However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step in the direction of a model-based, and thus scientific, analysis. However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series. Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented. Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.

@@

<!-- JR: Begin by defining, broadly, what Bayesian statistics is? -->

At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.). Our assumptions about the demographic development of the past must naturally be very conservative. In the logic of the basic Bayesian equation, these assumptions represent the prior (probability). In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian learning from data. It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction. Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.

@@

```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau."}
fc <- sf::st_read("../data/fixed_online_data/BiogeographischeRegionen.gdb", layer = "N2020_Revision_BiogeoRegion")
fc <- st_transform(fc, 4326)
fc <- fc[fc$RegionNummer==2,] %>% st_union()

main_map <- my_basemap +
  geom_sf(data = fc, inherit.aes = FALSE, alpha = 0.33, fill="darkred", color=NA) +
  blank() +
  my_scalebar

inset_map <- europe_overview_map +
  geom_sf(data = switzerland_bb, fill = NA, color = "darkred", inherit.aes = FALSE) +
  blank()

ggdraw() +
  draw_plot(main_map) +
  draw_plot(inset_map, x = 0, y = 0.7, width = 0.3, height = 0.3)
```

The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat but largely hilly area between the Jura Mountains and the Alps. Its average altitude above sea level is between 400 and 600 m. Due to its situation as a basin, it represents a favourable area and is by far the most densely populated region of the Switzerland today. Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape. The Swiss Plateau was shaped by glaciers during the Ice Age, which led to the many lakes that characterise the landscape today, and which provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements.

With a mild climate compared to the Jura and the Alps and a dense network of water bodies, the Swiss Plateau is a favourable area for human settlement in Switzerland. From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers on the one hand excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.

@@

@@

```{r compdendroc14, fig.cap="Comparison of the scaled (non-centered z-transformed) number of ^14^C and dendrodated sites over time in the dataset used."}
joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))

@@

@@

<!-- But in the data table that accompanies the repository, they would be available -->

@@

Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory. Formation conditions, in this case most likely C-transformation [@schiffer1987], prevent their use in this study. Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.

@@

In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods, i.e. the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on. In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time. This results in the following very simple formula:

@@

Thus, we can define the limits, the $max\_change\_rate$<!-- JR: no Greek letter? :( MH: I could invent one, if you like? ;-) --> as an time independent variable, again without already specifying them with fixed values at this stage. The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively. Overall, this represents a hierarchical model that can be noted as follows:

@@

The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$. This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model. The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:
